### 2017-03-20 main.tex
Add data and graph for new accuracy measurements.
Makefile: new target to build the diff (makes some unwarranted assumptions!)

% Use the asterisk to denote corresponding authorship and provide email address in note below.
\textsuperscript{\textcurrency} Corresponding author: bwpearre@gmail.com (BP)

@@

\begin{abstract}
  The song of the adult male zebra finch is strikingly stereotyped.
  Efforts to understand motor output, pattern generation, and learning
  have taken advantage of this consistency by investigating the bird's
  ability to modify specific parts of song under external cues, and by
  examining timing relationships between neural activity and vocal
  output.  Such experiments require that precise moments during song
  be identified in real time as the bird sings.  Various
  syllable-detection methods exist, but many require special hardware,
  software, and know-how, and details on their implementation and
  performance are scarce.  We present an accurate, versatile, and fast
  syllable detector that can control hardware at precisely timed
  moments during zebra finch song.  Many moments during song can be
  isolated and detected with false negative and false positive rates
  well under 1\% and 0.001\% respectively. The detector can run on a
  stock Mac Mini with triggering delay of less than a millisecond and
  a jitter of $\sigma\approx 2$ milliseconds.
\end{abstract}

@@

\begin{enumerate}
\item Record and align a corpus of training songs.  The technique has been published in \cite{Poole2012}. 200 songs generally yield acceptable results, but here we standardise on 1000-song training sets.
\item Choose one or more instants in the song that should create a trigger event, and train a
  neural network to recognise them. This step is carried out offline.  While any neural network software would produce similar results, we used MATLAB 2015b's neural network toolbox.
\item Once trained and saved, the neural network is used by a realtime
  detection programme that listens to an audio signal and indicates detection of the target
  syllables via a TTL pulse.  We present three detection implementations, in MATLAB, LabVIEW, and Swift, that trade off hardware requirements, ease of maintenance, and performance.
\end{enumerate}
This method makes the following contributions:
\begin{itemize}
\item Fast: sub-millisecond latencies, with jitter around 2 ms.
\item Accurate: false negative rates under 1\% and false positive rates under 0.001\% for a variety of syllables.
\item State-of-the-art performance with default parameters.
\item Requires almost no programming experience.
\item Runs on inexpensive hardware.
\item Described in detail here, with reference implementations provided and benchmarked.
\end{itemize}

@@

\begin{figure}
  % Width: 8, Height: 3, Fonts scale: 120%
  \includegraphics[width=\textwidth]{Fig1}
  \caption{The spectrogram of a typical song, used as the example
    throughout this paper.  This image was made by superposing the
    spectra of our 2818 aligned songs.  Our example detection points,
    $t^*_1\ldots t^*_6$, are shown as red lines, with the recognition
    regions (30 ms $\times$ 1--8 kHz) marked as rectangles.}
  \label{fig:song}
\end{figure}

Note that this results in time being discretised into chunks of length
$t_\mathrm{fft}$.  Because each Fourier transform computation contains
a small number $n_\mathrm{fft}$ of new audio samples (in the example
above, 40 new samples and $256-40$ that have already been examined),
we tried implementing the Sliding Discrete Fourier transform (SDFT)
\cite{Jacobsen2003SlidingDFT}.  This allows
$t_\textrm{fft}=t_\textrm{sample}$.  Practically, the operating system
retrieves new audio samples from the hardware several at a time, so
the full benefit of the SDFT is difficult to see in practice.
Furthermore, we found that FFT implementations are sufficiently highly
optimised that discretising time into chunks of $t_\mathrm{fft}$ as we
have done produced similar results with simpler software.

@@

The neural network's inputs are the FFT values from a rectangular
region of the spectrogram covering a predefined range of frequency
values $F$ (for example, 1--8 kHz) at some number of the most recent
frames $n_\textrm{fft}$.  Any time $t$ falls within frame
$\tau(t)$, and $t-t_\textrm{fft}$ falls within the previous frame, so the recognition region that the neural network receives as input
consists of the spectrogram values over $F$ at $\tau(t)$ and those
from the contiguous set of recent frames: $T = \{$ $\tau(t)$,
$\tau(t-t_\textrm{fft})$, $\tau(t-{2t_\textrm{fft}})\ldots$
$\tau(t-n_\textrm{fft}t_\textrm{fft})\}$.  We have found that a 30-ms
time window---which will yield $n_\textrm{fft} = |T|=\lfloor 30\textrm{
  ms}/t_\textrm{fft}\rceil$ frames---of frequencies spanning 1--8 kHz
works well for most syllables.

Six examples of chosen target moments in the song, and their
corresponding recognition regions $F$ and $T$, are shown in
\fig{fig:song}.

@@

We found these parameters to work well across a variety of target
syllables, but various other parameter sets yield results similar to
those presented here.  Some of the parameters trade off detection
accuracy or temporal precision vs.~training time. For example,
decreasing the frame interval generally decreases both latency and
jitter, but also increases training time.

@@

\fig{fig:song} shows a typical song. The image is produced by
averaging the aligned spectrograms for the bird lny64.  Six target
trigger points are shown for this song, at 150, 200, 250, 300, 350,
and 400 milliseconds after the beginning of the aligned samples, as
indicated by the red lines, and for convenience will be referred to
henceforth as $t^*_1\ldots t^*_6$ respectively.  For clarity, the
rectangles show a shorter recognition window than our standard: here
30 ms, but still spanning 1--8 kHz, for each target time in the
song.\marginpar{Is this OK?  Should I update to a 50-ms window?
  They'll overlap.}

@@

Accuracies for our six test syllables for bird lny64 are shown
in\marginpar{Update if I include all birds in this figure!}
\fig{fig:accuracies} (accuracies for the synthetic $\delta$-function
songs were always 100\%).  For each test syllable we trained 100
different networks, which differ from each other both in the random
initialisation of network weights and in which random subset of the
song-and-nonsong corpus was used for training.  Each network's
accuracy is shown as a single point in each chart in
\fig{fig:accuracies}, and the means are shown in
Table~\ref{table:accuracies}.  Evidently some syllables are easier to
identify reliably than others (for example, $t^*_3$ and $t^*_6$ are
occasionally confused due to their similar appearance, as could be
guessed from \fig{fig:song}); thus, if the experimental design
permits, it can be worthwhile to choose an easily identifiable
syllable.  Furthermore, as can be seen from \fig{fig:accuracies}, for
difficult syllables the training process occasionally yields a network
that performs unusually poorly.  It is easy to spot these networks by
their performance either as measured as part of the training process
or on the test audio file, perhaps taking the best of a few training
runs.

@@

  \caption{Accuracy variability over 100 different training runs for
    each of the six test syllables for lny64.  Each dot shows the test-set
    accuracy for an independently trained detector.  Because the
    horizontal positions have been randomised slightly so as not to
    occlude same-valued measurements, test syllable is also indicated
    by colour.  The means for all birds are given in Table~\ref{table:accuracies}.}
  \label{fig:accuracies}
\end{figure}

\begin{table}
  \begin{tabular}{rc|rr}
    Bird & Target time & \% True positives & \% False positives \\ \hline
    A&$ t^*_1$ & 99.72 &  0.0025 \\ % lny64
    &$ t^*_2$ & 99.81 &  0.00053 \\
    &$ t^*_3$ & 99.78 &  0.00062 \\
    &$ t^*_4$ & 99.69 &  0.00082 \\
    &$ t^*_5$ & 99.75 &  0.00028 \\
    &$ t^*_6$ & 99.54 &  0.0012 \\
    B&$ t^*$ & 99.370992640000068   &0.003887539045000 \\ % lny46
    C&$ t^*$ & 99.218862389999998   &0.001205787775200 \\ % lny42
    D&$ t^*$ & 99.342842499999975   &0.00260723410200  \\ % lny4rb
    E& $t^*$ & 99.726635259999981   &0.003169430259900 \\ % lr28
    E& $t^*$ & 99.464679732673261   &0.004515502156436 \\ % lr12
    E& $t^*$ & 99.862406879999966   &0.001215710723800 \\ % lr13
    E& $t^*$ & 99.646613329999980   &0.003110002775000 % lr77
  \end{tabular}
  \caption{Mean values for the detection accuracies shown in \fig{fig:accuracies}.}
  \label{table:accuracies}
\end{table}

@@

@@

We compared latency and jitter across our different detector
implementations for the $\delta$-syllable and lny64's $t^*_4$, again with
$t_\mathrm{fft}=1.5$ ms.  Results are as follows:
\vspace{8pt}
\begin{tabular}{l|c|c|c| c}
  & \multicolumn{2}{c}{$\delta$-syllable} & \multicolumn{2}{c}{Bird lny64: $t^*_4$} \\
  Detector & Latency (ms) & Jitter (ms) & Latency (ms) & Jitter (ms) \\
  \hline
  Ideal & 0.66 & 0.38 & -1.0 & 2.0 \\
  LabVIEW & 1.5 & 0.69 & -0.18 & 2.4 \\
  Swift+serial & 0.88 & 1.6 & -0.075 & 2.0 \\
  Swift+audio & 5.9 & 2.6 & 1.2 & 4.9 \\
  MATLAB+serial & 8.1 & 2.3 & 3.8 & 2.0 \\
  MATLAB+audio & 13 & 5.9 & 8.4 & 5.9
\end{tabular}
\vspace{8pt}\par\noindent
These data are also plotted in \fig{fig:TimingVsDetector}, and
\fig{fig:timing} gives a more detailed view of what the timing curves
look like for the five implementations of our detector on $t^*_4$.

@@

This syllable detector is appropriate for zebra finch song, and
although our tests were carried out on songs from that species, it is
also likely to work well for Bengalese finches due to their similarly stereotyped syllables.  It offers the
following benefits:
\begin{itemize}
\item The detector is accurate. False negative and false positive
  rates can be well under 1\% and 0.001\% respectively, and trading
  these two numbers off against each other is through a single
  relative-cost parameter.
\item Latency is generally under a millisecond, with jitter around 2
  ms.
\item Works on a wide range of target syllables using the default
  values described here, generally eliminating the need for
  hand-tuning.
\item Runs fast enough for syllable-modification experiments on
  inexpensive consumer-grade hardware, although we recommend that the
  training phase be run on a fast desktop system with 32 GB of RAM.
\item A single detector can generate different target pulses for
  multiple syllables at almost no additional computational cost during
  runtime, although training time will increase.
\end{itemize}

>>>>

% Use the asterisk to denote corresponding authorship and provide email address in note below.
\textsuperscript{\textcurrency} Corresponding author: bwpearre@bu.edu (BP)

@@

\begin{abstract}
  The song of the adult male zebra finch is strikingly stereotyped.
  Efforts to understand motor output, pattern generation, and learning
  have taken advantage of this consistency by investigating the bird's
  ability to modify specific parts of song under external cues, and by
  examining timing relationships between neural activity and vocal
  output.  Such experiments require that precise moments during song
  be identified in real time as the bird sings.  Various
  syllable-detection methods exist, but many require special hardware,
  software, and know-how, and details on their implementation and
  performance are scarce.  We present an accurate, versatile, and fast
  syllable detector that can control hardware at precisely timed
  moments during zebra finch song.  Many moments during song can be
  isolated and detected with false negative and false positive rates
  well under 1\% and 0.005\% respectively. The detector can run on a
  stock Mac Mini with triggering delay of less than a millisecond and
  a jitter of $\sigma\approx 2$ milliseconds.
\end{abstract}

@@

\begin{enumerate}
\item Record and align a corpus of training songs.  The technique has been published in \cite{Poole2012}. 200 songs generally yield acceptable results, but here we standardise on 1000-song training sets.
\item Choose one or more instants in the song that should create a trigger event, and train a
  neural network to recognise them. This step is carried out offline.  While any neural network software would produce similar results, we used MATLAB 2015b's neural network toolbox.
\item Once trained and saved, the neural network is used by a realtime
  detection programme that listens to an audio signal and indicates detection of the target
  syllables via a TTL pulse.  We present three detection implementations, in MATLAB, LabVIEW, and Swift, that trade off hardware requirements, ease of maintenance, and performance.
\end{enumerate}
This method makes the following contributions:
\begin{itemize}
\item Fast: sub-millisecond latencies, with jitter around 2 ms.
\item Accurate: false negative rates under 1\% and false positive rates under 0.005\% for a variety of syllables.
\item State-of-the-art performance with default parameters.
\item Requires almost no programming experience.
\item Runs on inexpensive hardware.
\item Described in detail here, with reference implementations provided and benchmarked.
\end{itemize}

@@

\begin{figure}
  % Width: 8, Height: 3, Fonts scale: 120%
  \includegraphics[width=\textwidth]{Fig1}
  \caption{The spectrogram of the song of lny64, used as an example
    throughout this paper.  This image was made by superposing the
    spectra of our 2818 aligned songs.  Our example detection points,
    $t^*_1\ldots t^*_6$, are shown as red lines, with example recognition
    regions of 30 ms $\times$ 1--8 kHz marked as rectangles.}
  \label{fig:song}
\end{figure}

This results in time being discretised into chunks of length
$t_\mathrm{fft}$.  Because each Fourier transform computation contains
a small number $n_\mathrm{fft}$ of new audio samples (in the example
above, 40 new samples and $256-40$ that have already been examined),
we tried implementing the Sliding Discrete Fourier transform (SDFT)
\cite{Jacobsen2003SlidingDFT}.  This allows
$t_\textrm{fft}=t_\textrm{sample}$.  Practically, the operating system
retrieves new audio samples from the hardware several at a time, so
the full benefit of the SDFT is difficult to see in practice.
Furthermore, we found that FFT implementations are sufficiently highly
optimised that discretising time into chunks of $t_\mathrm{fft}$ as we
have done produced similar results with simpler software.

@@

The neural network's inputs are the FFT values from a rectangular
region of the spectrogram covering a predefined range of frequency
values $F$ (for example, 1--8 kHz) at some number of the most recent
frames $n_\textrm{fft}$.  Any time $t$ falls within frame $\tau(t)$,
and $t-t_\textrm{fft}$ falls within the previous frame, so the
recognition region that the neural network receives as input consists
of the spectrogram values over $F$ at $\tau(t)$ and those from the
contiguous set of recent frames: $T = \{$ $\tau(t)$,
$\tau(t-t_\textrm{fft})$, $\tau(t-{2t_\textrm{fft}})\ldots$
$\tau(t-n_\textrm{fft}t_\textrm{fft})\}$.  Time windows of 30--50
ms---the latter will yield $n_\textrm{fft} = |T|=\lfloor 50\textrm{
  ms}/t_\textrm{fft}\rceil$ frames---of frequencies spanning 1--8 kHz
generally work well.

Six examples of chosen target moments in the song, with recognition
regions $F = 1--8$ kHz and $T=30$ ms, are shown in \fig{fig:song}.

@@

We found these parameters to work well across a variety of target
syllables, but various other parameter sets yield results similar to
those presented here.  Some of the parameters trade off detection
accuracy or temporal precision vs.~training time. For example,
decreasing the frame interval generally decreases both latency and
jitter, but also increases training time. Sometimes the effects are
syllable-specific: for example, using a 30-ms time window rather than
50 ms speeds training while usually having a minimal effect on
detector performance, but occasionally a syllable will be seen for
which extending the window to 80 ms is very helpful.

@@

\fig{fig:song} shows a typical song. The image is produced by
averaging the aligned spectrograms for the bird lny64.  Six target
trigger points are shown for this song, at 150, 200, 250, 300, 350,
and 400 milliseconds after the beginning of the aligned samples, as
indicated by the red lines. In discussion, we will refer to these six
as $t^*_1\ldots t^*_6$ respectively.  For clarity, the rectangles show
a shorter recognition window than our standard: here 30 ms, but still
spanning 1--8 kHz, for each target time in the song.\marginpar{Is this
  OK?  Should I update to a 50-ms window?  They'll overlap.}

@@

Accuracies are shown in \fig{fig:accuracies} (accuracies for the synthetic $\delta$-function
songs were always 100\%).  For each test syllable we trained 100
different networks, which differ from each other both in the random
initialisation of network weights and in the random subset of the
song-and-nonsong corpus that was used for training.  Each network's
accuracy is shown as a single point in each chart in
\fig{fig:accuracies}, and the means are shown in
Table~\ref{table:accuracies}.  Evidently some syllables are easier to
identify reliably than others (for example, for lny64, $t^*_3$ (250 ms) and $t^*_6$ (400 ms) are
occasionally confused due to their similar appearance, as could be
guessed from \fig{fig:song}); thus, if the experimental design
permits, it can be worthwhile to choose an easily identifiable
syllable.  Furthermore, as can be seen from \fig{fig:accuracies}, for
difficult syllables the training process occasionally yields a network
that performs unusually poorly.  It is easy to spot these networks by
their performance either as measured as part of the training process
or on the test audio file, perhaps taking the best of a few training
runs.

@@

  \caption{Accuracy variability over 100 different training runs for
    each of the test detection points.  Each dot shows the test-set
    accuracy for an independently trained detector.  Because the
    horizontal positions have been randomised slightly so as not to
    occlude same-valued measurements, test syllable is also indicated
    by colour.  The means are given in Table~\ref{table:accuracies}.}
  \label{fig:accuracies}
\end{figure}

\begin{table}
  % replot_accuracies_concatanated('tex', true) for confusion_log_perf.txt (1.5 ms).
  \begin{tabular}{lc|cc}
    Bird & Target time (ms) & \% True positives & \% False positives \\ \hline
lny64 & 	150 $[t^*_1]$ & 	99.66 & 	0.0026 \\ % /Volumes/Data/song/lny64/confusion_log_perf.txt
      & 	200 $[t^*_2]$ & 	99.75 & 	0.00052 \\ %                                               
      & 	250 $[t^*_3]$ & 	99.79 & 	0.00051 \\ %                                               
      & 	300 $[t^*_4]$ & 	99.70 & 	0.00077 \\ %                                               
      & 	350 $[t^*_5]$ & 	99.73 & 	0.00022 \\ %                                               
      & 	400 $[t^*_6]$ & 	99.52 & 	0.0010 \\ %                                                
lny42 & 	380 & 	99.09 & 	0.0011 \\ % /Volumes/Data/song/lny42/confusion_log_perf.txt          
lny46 & 	140 & 	99.53 & 	0.0041 \\ % /Volumes/Data/song/lny46/confusion_log_perf.txt          
lny4rb & 	250 & 	99.36 & 	0.0025 \\ % /Volumes/Data/song/lny4rb/confusion_log_perf.txt        
lr12 & 	180 & 	99.52 & 	0.0033 \\ % /Volumes/Data/song/lr12/all/confusion_log_perf.txt        
lr13 & 	300 & 	99.83 & 	0.0013 \\ % /Volumes/Data/song/lr13/all/confusion_log_perf.txt        
lr28 & 	140 & 	99.73 & 	0.0028 \\ % /Volumes/Data/song/lr28/all/confusion_log_perf.txt        
lr77 & 	320 & 	99.58 & 	0.0028  % /Volumes/Data/song/lr77/all/confusion_log_perf.txt          
  \end{tabular}
  \caption{Mean values for the detection accuracies shown in \fig{fig:accuracies}.}
  \label{table:accuracies}
\end{table}

@@

Changing the FFT interval to 2 ms significantly reduces the
computational requirement. We found that some syllables were slightly
easier to detect and some more difficult, with no significant
difference on average.

@@

We compared latency and jitter across our different detector
implementations for the $\delta$-syllable and lny64's $t^*_4$, again with
$t_\mathrm{fft}=1.5$ ms.  Results are as follows:
\vspace{8pt}
\begin{tabular}{l|c|c|c|c}
  & \multicolumn{2}{c}{$\delta$-syllable} & \multicolumn{2}{c}{Bird: $t^*_4$} \\
  Detector & Latency (ms) & Jitter (ms) & Latency (ms) & Jitter (ms) \\
  \hline
  Ideal & 0.66 & 0.38 & -1.0 & 2.0 \\
  LabVIEW & 1.5 & 0.69 & -0.18 & 2.4 \\
  Swift+serial & 0.88 & 1.6 & -0.075 & 2.0 \\
  Swift+audio & 5.9 & 2.6 & 1.2 & 4.9 \\
  MATLAB+serial & 8.1 & 2.3 & 3.8 & 2.0 \\
  MATLAB+audio & 13 & 5.9 & 8.4 & 5.9
\end{tabular}
\vspace{8pt}\par\noindent
These data are also plotted in \fig{fig:TimingVsDetector}, and
\fig{fig:timing} gives a more detailed view of what the timing curves
look like for the five implementations of our detector on $t^*_4$.

@@

This syllable detector is appropriate for zebra finch song, and
although our tests were carried out on songs from that species, it is
also likely to work well for Bengalese finches due to their similarly stereotyped syllables.  It offers the
following benefits:
\begin{itemize}
\item The detector is accurate. False negative and false positive
  rates can be well under 1\% and 0.005\% respectively, and trading
  these two numbers off against each other is through a single
  relative-cost parameter.
\item Latency is generally under a millisecond, with jitter around 2
  ms.
\item Works on a wide range of target syllables using the default
  values described here, generally eliminating the need for
  hand-tuning.
\item Runs fast enough for syllable-modification experiments on
  inexpensive consumer-grade hardware, although we recommend that the
  training phase be run on a fast desktop system with 32 GB of RAM.
\item A single detector can generate different target pulses for
  multiple syllables at almost no additional computational cost during
  runtime, although training time will increase.
\end{itemize}
